---
title: "Malcolm Newell Reds Code"
author: "Malcolm Newell"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path[1]))
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
```

### Libraries 
```{r libraries, message=FALSE}
suppressWarnings(suppressMessages({
  library(dplyr)
  library(tidyr)
  library(readr)
  library(mgcv)
  library(ggplot2)
  library(ggeasy)
}))
```

### Functions
```{r functions, message=FALSE}
get_gam_model <- function(train_data){
  
  model <- gam(usage_percent ~ 
                 s(DELTA_RUN_EXP, BALLS) + 
                 s(DELTA_RUN_EXP, STRIKES) +
                 as.factor(BAT_SIDE) + as.factor(THROW_SIDE),
                      data = train_data, method = "REML")
  
}

# Get Player Search Data
get_search_data <- function(player_name){
  
  df <- search_joined_data %>%
    filter(PLAYER_NAME == "player_name") %>%
    arrange(desc(xwOBA)) %>%
    select("Side" = THROW_SIDE, "Pitch Group" = pitch_group,
           "Usage %" = usage_percent, `Max EV`, `Avg EV`, LA, xBA, xwOBA, wOBA)
  
}

get_prediction_plot <- function(df_predictions){
  
  data_long <- df_predictions %>% 
  dplyr::rename(
    Fastball = PITCH_TYPE_FB,
    "Breaking Ball" = PITCH_TYPE_BB,
    Offspeed = PITCH_TYPE_OS
  ) %>%
  tidyr::pivot_longer(cols = c(Fastball, `Breaking Ball`, Offspeed),
                      names_to = "Pitch Group",
                      values_to = "Usage_Percentage") %>%
  dplyr::mutate(
    `Pitch Group` = factor(`Pitch Group`, 
                           levels = c("Fastball", "Breaking Ball", "Offspeed"))
  )
  

# Create bar plot
ggplot(data_long, aes(x = `Pitch Group`, y = Usage_Percentage, fill = `Pitch Group`)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(Usage_Percentage, "%")), 
            position = position_stack(vjust = 0.5), 
            color = "white") +
  labs(title = paste0("Pitch Usage Percentage for ", unique(data_long$PLAYER_NAME), " in 2024"),
       x = "",
       y = "Usage %") +
  theme_minimal() +
  ggeasy::easy_center_title() +
  ggeasy::easy_remove_legend() +
  scale_fill_manual(values = c("Fastball" = "#C6011F", 
                                "Breaking Ball" = "skyblue", 
                                "Offspeed" = "darkgreen")) +
  theme(panel.grid.major = element_blank()) +
  ylim(0, 50)

}

```


### Reading Data

```{r data, message=FALSE}
data <- read_csv(file = "Data/data.csv")
predictions <- read_csv(file = "Data/predictions.csv")
sample <- read_csv(file = "Data/sample_submission.csv")
```


### Changing pitch types to pitch groups
#### I did not include the Ephus, Knuckleball, Other, or Pitch Out in the groups
```{r pitch_group, message=FALSE}
pitch_data <- data %>%
  dplyr::mutate(
    pitch_id = row_number(),
    pitch_group = ifelse(
    PITCH_TYPE %in% c("FF", "SI"), "FB",
    ifelse(
      PITCH_TYPE %in% c("CH", "FO", "FS", "SC"), "OS",
      ifelse(
        PITCH_TYPE %in% c("CS", "CU", "FC", "KC", "SL", "ST", "SV"), "BB", 
        "Other"
      )
    )
  )) %>%
  dplyr::filter(pitch_group != "Other") %>%
  select(pitch_id, BATTER_ID, PLAYER_NAME, pitch_group,
         BAT_SIDE, THROW_SIDE,
         BALLS, STRIKES, DELTA_RUN_EXP)

```

### Finding Usage % for each pitch group and player every year
```{r usage, message=FALSE}
usages <- pitch_data %>%
  group_by(BATTER_ID, PLAYER_NAME, THROW_SIDE) %>%
  dplyr::mutate(total_pitches = n()) %>%
  ungroup() %>%
  group_by(BATTER_ID, PLAYER_NAME, THROW_SIDE, pitch_group) %>%
  dplyr::summarize(group_pitches = n(),
                   total_pitches = first(total_pitches)) %>%
  dplyr::mutate(usage_percent = (group_pitches / total_pitches) * 100) %>%
  ungroup()

```

### Combine pitch_data with usages
```{r joining, message=FALSE}
joined_data <- usages %>%
  dplyr::left_join(pitch_data, by = c("BATTER_ID", "PLAYER_NAME",
                                      "THROW_SIDE", "pitch_group"))

```


### Set the seed and prep the model by filtering dataframes for each pitch group
```{r prep, message=FALSE}
addTaskCallback(function(...){set.seed(123);TRUE})

fb_data <- joined_data %>% dplyr::filter(pitch_group == "FB")
bb_data <- joined_data %>% dplyr::filter(pitch_group == "BB")
os_data <- joined_data %>% dplyr::filter(pitch_group == "OS")

```
## Fastball Model
```{r fastball}
dt_fb <- sample(nrow(fb_data), nrow(fb_data) * .7)
train_fb <- fb_data[dt_fb,]
test_fb <- fb_data[-dt_fb,]

fb_usage_model <- get_gam_model(train_fb)

summary(fb_usage_model)

```

## Breaking Ball Model
```{r breaking_ball}
dt_bb <- sample(nrow(bb_data), nrow(bb_data) * .7)
train_bb <- bb_data[dt_bb,]
test_bb <- bb_data[-dt_bb,]

bb_usage_model <- get_gam_model(train_bb)

summary(bb_usage_model)

```

## Off-speed Model
```{r offspeed}
dt_os <- sample(nrow(os_data), nrow(os_data) * .7)
train_os <- os_data[dt_os,]
test_os <- os_data[-dt_os,]

os_usage_model <- get_gam_model(train_os)

summary(os_usage_model)

```

## Predictions
```{r predictions, message=FALSE}
# Predict fastballs
predictions_fb <- test_fb %>%
  dplyr::mutate(GAME_YEAR = 2024)

predictions_fb$fastball_usage <- predict(fb_usage_model, newdata =
                                           predictions_fb, type = "response")

predictions_fb <- predictions_fb %>%
  dplyr::select(pitch_id, BATTER_ID, PLAYER_NAME, GAME_YEAR, 
                THROW_SIDE, fastball_usage)

# Predict breaking balls
predictions_bb <- test_bb %>%
  mutate(GAME_YEAR = 2024)

predictions_bb$breaking_usage <- predict(bb_usage_model, newdata =
                                           predictions_bb, type = "response")

predictions_bb <- predictions_bb %>%
  dplyr::select(pitch_id, BATTER_ID, PLAYER_NAME, GAME_YEAR, 
                THROW_SIDE, breaking_usage)

# Predict off-speed
predictions_os <- test_os %>%
  mutate(GAME_YEAR = 2024)

predictions_os$offspeed_usage <- predict(os_usage_model, newdata =
                                           predictions_os, type = "response")
predictions_os <- predictions_os %>%
  dplyr::select(pitch_id, BATTER_ID, PLAYER_NAME, GAME_YEAR, 
                THROW_SIDE, offspeed_usage)

```

## Combining Predictions into one dataframe
#### Joining all predictions then aggregating the results
```{r combined_predictions}

joined_predictions <- predictions_fb %>%
  dplyr::full_join(
    predictions_bb, by = c("pitch_id", "BATTER_ID", "PLAYER_NAME", 
                           "GAME_YEAR", "THROW_SIDE")
  ) %>%
  dplyr::full_join(
    predictions_os, by = c("pitch_id", "BATTER_ID", "PLAYER_NAME", 
                           "GAME_YEAR", "THROW_SIDE")
  )

predictions <- joined_predictions %>%
  group_by(BATTER_ID, PLAYER_NAME, GAME_YEAR) %>%
  dplyr::summarize(
    PITCH_TYPE_FB = round(mean(fastball_usage, na.rm = TRUE),1),
    PITCH_TYPE_BB = round(mean(breaking_usage, na.rm = TRUE),1),
    PITCH_TYPE_OS = round(mean(offspeed_usage, na.rm = TRUE),1)
  ) %>%
  ungroup() %>%
  dplyr::select(BATTER_ID, PLAYER_NAME, GAME_YEAR, PITCH_TYPE_FB,
                PITCH_TYPE_BB, PITCH_TYPE_OS)

```
## Graphics
#### Using this data to find three interesting players
```{r search, message=FALSE}

search_pitch_data <- data %>%
  dplyr::mutate(
    pitch_id = row_number(),
    pitch_group = ifelse(
    PITCH_TYPE %in% c("FF", "SI"), "FB",
    ifelse(
      PITCH_TYPE %in% c("CH", "FO", "FS", "SC"), "OS",
      ifelse(
        PITCH_TYPE %in% c("CS", "CU", "FC", "KC", "SL", "ST", "SV"), "BB", 
        "Other"
      )
    )
  )) %>%
  dplyr::filter(pitch_group != "Other") 

aggregate_data <- search_pitch_data %>%
  group_by(BATTER_ID, PLAYER_NAME,  THROW_SIDE, pitch_group) %>%
  dplyr::summarize(
    "Max EV" = round(max(LAUNCH_SPEED, na.rm = T),1),
    "Avg EV" = round(mean(LAUNCH_SPEED, na.rm = T),1),
    LA = round(mean(LAUNCH_ANGLE, na.rm = T)),
    xBA = round(mean(ESTIMATED_BA_USING_SPEEDANGLE, na.rm = T),3),
    xwOBA = round(mean(ESTIMATED_WOBA_USING_SPEEDANGLE, na.rm = T),3),
    wOBA = round(mean(WOBA_VALUE, na.rm = T),3)
  ) %>%
  ungroup()
  

search_usages <- search_pitch_data %>%
  group_by(BATTER_ID, PLAYER_NAME, THROW_SIDE) %>%
  dplyr::mutate(total_pitches = n()) %>%
  ungroup() %>%
  group_by(BATTER_ID, PLAYER_NAME, THROW_SIDE, pitch_group) %>%
  dplyr::summarize(group_pitches = n(),
                   total_pitches = first(total_pitches)) %>%
  dplyr::mutate(usage_percent = round((group_pitches / 
                                         total_pitches) * 100, 1)) %>%
  ungroup()

search_joined_data <- search_usages %>%
  dplyr::left_join(aggregate_data, by = c("BATTER_ID", "PLAYER_NAME",
                                      "THROW_SIDE", "pitch_group"))

```

### Individual Player Metrics
```{r metrics, message=FALSE}

nimmo_df <- get_search_data("Nimmo, Brandon")

teoscar_df <- get_search_data("Hernández, Teoscar")

steer_df <- get_search_data("Steer, Spencer")

```

## View Prediction Plots
```{r plots}

nimmo_predictions <- predictions %>%
  filter(PLAYER_NAME == "Nimmo, Brandon")

teoscar_predictions <- predictions %>%
  filter(PLAYER_NAME == "Hernández, Teoscar")

steer_predictions <- predictions %>%
  filter(PLAYER_NAME == "Steer, Spencer")

get_prediction_plot(nimmo_predictions)

get_prediction_plot(teoscar_predictions)

get_prediction_plot(steer_predictions)

```

